### 2016-03-15 paper.tex
Add Wosnik edits to introduction

The Reference Model Project (RMP), sponsored by the US Department of Energy,
produced six marine hydrokinetic (MHK) technology point designs as reference
models (RMs) to serve as non-proprietary test articles for open research
development, and to benchmark performance and costs for technology developers
\cite{Neary2014, Neary2014a}. Open-source RMP products, along with supporting
documentation, are available at the RMP website
\url{http://energy.sandia.gov/rmp} to facilitate their use in future R\&D
studies by industry, academia, and national laboratories. These products
include: Technical specifications and computer-aided design (CAD) files for each
RM device to allow exact replication for physical and numerical modeling
studies; resource site information used to design each RM device; and references
to physical modeling data sets that can be used to validate numerical modeling
design and analysis tools.

Reference Model 2 (RM2) is a dual-rotor, vertical-axis cross-flow river turbine
that was designed to operate in a reach of the lower Mississippi River near
Baton Rouge, Louisiana \cite{Barone2011, Neary2011}. The rotor has three tapered
blades and a relatively low solidity $Nc/(\pi D)$ or chord-to-radius ratio $c/R
\approx 0.1$. A preliminary analysis with Sandia's CACTUS vortex line numerical
model \cite{Murray2011} predicted an individual rotor's full-scale performance
coefficient to be 47\% at 1 m/s flow speed and a tip speed ratio $\lambda \equiv
\frac{\omega R}{U_\infty} = 3.15$ \cite{Barone2011}, where $\omega$ and $R$ are
the rotor's angular velocity and radius, respectively. Note that the RM2 rotor
solidity would be considered moderate to high for a wind turbine, but MHK rotors
are typically higher solidity since they must withstand approximately an order
of magnitude higher torque in typical flows, necessitating relatively larger
blades to meet strength and fatigue life requirements.

An initial experimental measurement campaign with a 1:15 scale RM2 rotor
conducted at the Saint Anthony Falls Laboratory (SAFL) at the University of
Minnesota resulted in maximum power coefficient of approximately 5\% at $\lambda
= 2.2$. This discrepancy between numerical and physical model performance
prompted the experimental measurements presented here, namely due to the low
Reynolds number---$Re_D = U_\infty D / \nu$, where $U_\infty$ is the free stream
velocity, $D$ is the rotor diameter, and $\nu$ is the fluid kinematic
viscosity---of the 1:15 scale tests, which were performed at $Re_D \sim 10^5$
\cite{Hill2014}.

The effect of Reynolds number on average power output was shown to be
significant on the 2 m Sandia Research Darrieus turbine in wind tunnel testing
\cite{Blackwell1976}: The maximum power coefficient, $C_{P_{\max}}$, increased
with Reynolds number, $Re_c$ (based on blade chord rather than diameter, though
these are approximately proportional), along with a shift of the location of
$C_{P_{\max}}$ toward lower tip speed ratios due to delayed blade stall. The
effects of Reynolds number were quite dramatic over a relatively small range of
$Re_c \approx 1.1 \times 10^5$--$2.9 \times 10^5$. More recently, Bachant and
Wosnik \cite{Bachant2014, Bachant2016-RVAT-Re-dep} showed that performance and
near-wake characteristics of a high solidity cross-flow turbine become Reynolds
number independent at $Re_D \approx 10^6$ or $Re_c \approx 2 \times 10^5$.

@@

The main objective of the present study was to acquire a new experimental
dataset for the RM2 turbine at sufficient Reynolds numbers to be relevant to
full-scale physical and numerical modeling. It was hypothesized that the
parasitic losses from rotor blade support struts could play an important role in
overall turbine performance, especially for a lower solidity turbine that
operates at higher tip speed ratios. Thus, the parasitic torque from strut drag
was measured without blades, then deliberately and significantly increased in
the physical model to provide data to investigate its importance. The velocity
field in the near-wake of the turbine was then measured to compare with
measurements from a higher solidity rotor in the same experimental setup
\cite{Bachant2015-JoT}, and to provide validation data for numerical models that
predict wake flows, which determine turbine--turbine interaction and optimal
spacing for turbine arrays. This dataset, along with the code for processing and
visualization, has been made publicly available (licensed via Creative Commons
CC-BY for data and MIT license for code) as a Git repository on GitHub
(\url{https://github.com/UNH-CORE/RM2-tow-tank}) and as a citable archive with a
digital object identifier (DOI) \cite{Bachant2015-RM2-data}.

@@

    \begin{tabular}{c|c|c|c|c|c}
        Name & Rotor type & $N_\mathrm{b}$ & $c/R$ & $Re_D$ & Used in \\ 
        \hline
        Sandia 2 m \cite{Blackwell1976} &  Darrieus  & 2--3 & 0.06--0.09 & $\sim 10^6$ & \cite{Roh2013,Bedon2014} \\ 
        Sandia 5 m \cite{Sheldahl1977} &  Darrieus  & 3 & 0.08 & $\sim 10^6$ & \cite{Antheaume2008,Bedon2014} \\ 
        Sandia 17 m \cite{Worstell1978} & Darrieus  & 2 & 0.06 & $\sim 10^7$ & \cite{Para1988,Orlandi2015,Bedon2014} \\ 
        Sandia 34 m \cite{Ashwill1992} & Darrieus  & 2 & 0.05 & $\sim 10^7$ & \cite{Liu1992,Murray2011,Bedon2014}  \\ 
        Strickland \emph{et al.} \cite{Strickland1981} & H & 1--3 & 0.15 & $\sim 10^5$ & \cite{Ponta2001,Scheurich2011b} \\ 
        Laneville and Vitecoq \cite{Laneville1986} & H & 2 & 0.13 & $\sim 10^6$ & \cite{Amet2009} \\ 
        Howell \emph{et al.} \cite{Howell2010} & H & 2--3 & 0.33 & $\sim 10^5$ & \cite{Joo2015} \\ 
        Mertens \cite{Mertens2003} & H & 2 & 0.21 & $\sim 10^5$ & \cite{Orlandi2015} \\ 
        VAWT 850 \cite{Mays1990} & Tapered H & 2 & 0.05 & $\sim 10^7$ & \cite{Murray2011} \\ 
        UNH-RVAT \cite{Bachant2014-RVAT-baseline} & H & 3 & 0.28 & $\sim 10^6$ & \cite{Michelen2014}
    \end{tabular}

@@

In general, there have been more experiments done with rotors of low $c/R$,
which makes them easier to model, since unsteady dynamic effects are less
influential on the overall performance \cite{Strickland1981}. This is apparent
when examining the effectiveness of numerical models that rely on static foil
coefficient input data, e.g., streamtube and vortex models, which are most
applicable for $c/R \leq 0.1$. For example, Bedon \emph{et al.}~\cite{Bedon2014}
used a double multiple streamtube momentum model without dynamic stall
corrections to evaluate the effectiveness of various foil coefficient databases
against the Sandia Darrieus turbine experimental data. Despite using such a
simple model, performance predictions were quite accurate in most conditions
except at low tip speed ratio for the 2 m turbine, which had the highest $c/R$
of all the Sandia rotors, making the dynamic effects more important. This point
highlights the need for more validation data for higher solidity rotors to
ensure numerical models are robust enough to explore unique cross-flow turbine
designs, especially as the MHK concepts mature.

The UNH-RVAT, for which the performance, near-wake, and Reynolds number
dependence of was investigated using the same experimental setup as the study
here \cite{Bachant2015-JoT, Bachant2016-Energies}, was an H-rotor of 1 m height
and 1 m diameter. The UNH-RVAT experimental datasets are also openly available
\cite{Bachant2014-RVAT-baseline, Bachant2016-RVAT-Re-dep}, and provide an
interesting comparison for the near-wake dynamics of a rotor similar in scale to
the 1:6 RM2, but with non-tapered blades and a high solidity $c/R = 0.28$.

>>>>

The Reference Model Project (RMP), sponsored by the US Department of Energy
(DOE), produced six marine hydrokinetic (MHK) technology point designs as
reference models (RMs) to serve as non-proprietary test articles for open
research development, and to benchmark performance and costs for technology
developers \cite{Neary2014, Neary2014a}. Open-source RMP products, along with
supporting documentation, are available at the RMP website
\url{http://energy.sandia.gov/rmp} to facilitate their use in future R\&D
studies by industry, academia, and national laboratories. These products
include: Technical specifications and computer-aided design (CAD) files for each
RM device to allow exact replication for physical and numerical modeling
studies; resource site information used to design each RM device; and references
to physical modeling data sets that can be used to validate numerical modeling
design and analysis tools.

Reference Model 2 (RM2) is a dual-rotor, vertical-axis cross-flow hydrokinetic
(river) turbine that was designed to operate in a reach of the lower Mississippi
River near Baton Rouge, Louisiana \cite{Barone2011, Neary2011}. The rotor has
three tapered blades and a relatively low solidity $Nc/(\pi D)$ or
chord-to-radius ratio $c/R \approx 0.1$. A preliminary analysis with Sandia's
CACTUS vortex line numerical model \cite{Murray2011} predicted an individual
rotor's full-scale performance coefficient to be $C_P=0.47$ at 1 m/s flow speed
and a tip speed ratio $\lambda \equiv \frac{\omega R}{U_\infty} = 3.15$
\cite{Barone2011}, where $\omega$ and $R$ are the rotor's angular velocity and
radius, respectively. Note that the RM2 rotor solidity would be considered
moderate to high for a wind turbine, whose $c/R$ values typically range from
0.05 to 0.09, but MHK rotors are typically higher solidity since they must
withstand approximately an order of magnitude higher torque in typical flows,
necessitating relatively larger blades to meet strength and fatigue life
requirements.

An initial experimental measurement campaign with a 1:15 scale RM2 rotor
conducted at the Saint Anthony Falls Laboratory (SAFL) at the University of
Minnesota resulted in maximum power coefficient of approximately 5\% at $\lambda
= 2.2$ \cite{Hill2014}. This discrepancy between numerical and physical model
performance prompted the experimental measurements presented here, namely due to
the low Reynolds number---$Re_D = U_\infty D / \nu$, where $U_\infty$ is the
free stream velocity, $D$ is the rotor diameter, and $\nu$ is the fluid
kinematic viscosity---of the 1:15 scale tests, which were performed at $Re_D
\sim 10^5$.

The effect of Reynolds number on average power output was shown to be
significant on the 2 m Sandia Research Darrieus turbine in wind tunnel testing
\cite{Blackwell1976}: The maximum power coefficient, $C_{P_{\max}}$, increased
with Reynolds number, $Re_c=\lambda U_\infty c / \nu$ (based on blade chord
rather than diameter, though these are approximately proportional since optimal
tip speed ratio correlates inversely with solitidy and therefore chord length),
along with a shift of the location of $C_{P_{\max}}$ toward lower tip speed
ratios due to delayed blade stall. The effects of Reynolds number were quite
dramatic over a relatively small range of $Re_c \approx 1.1 \times 10^5$--$2.9
\times 10^5$. More recently, Bachant and Wosnik \cite{Bachant2014,
    Bachant2016-RVAT-Re-dep} showed that performance and near-wake characteristics
of a high solidity cross-flow turbine become Reynolds number independent at
$Re_D \approx 10^6$ or $Re_c \approx 2 \times 10^5$.

@@

The main objective of the present study was to acquire a new experimental
dataset for the RM2 turbine at sufficiently high Reynolds numbers to be relevant
to full-scale physical and numerical modeling. It was hypothesized that the
parasitic losses from rotor blade support struts could play an important role in
overall turbine performance, especially for a lower solidity turbine that
operates at higher tip speed ratios. Thus, the parasitic torque from strut drag
was measured without blades, then deliberately and significantly increased in
the physical model to provide data to investigate its importance. The velocity
field in the near-wake of the turbine was then measured to compare with
measurements from a higher solidity rotor in the same experimental setup
\cite{Bachant2015-JoT}, and to provide validation data for numerical models that
predict wake flows, which determine turbine--turbine interaction and optimal
spacing for turbine arrays. This dataset, along with the code for processing and
visualization, has been made publicly available (licensed via Creative Commons
CC-BY for data and MIT license for code) as a Git repository on GitHub
(\url{https://github.com/UNH-CORE/RM2-tow-tank}) and as a citable archive with a
digital object identifier (DOI) \cite{Bachant2015-RM2-data}.

@@

    \begin{tabular}{c|c|c|c|c|c}
        Name & Rotor type & $N_\mathrm{b}$ & $c/R$ & $Re_D$ & Used in \\ 
        \hline
        Sandia 2 m \cite{Blackwell1976} &  Darrieus  & 2--3 & 0.06--0.09 & $\sim 10^6$ & \cite{Roh2013,Bedon2014} \\ 
        Sandia 5 m \cite{Sheldahl1977} &  Darrieus  & 3 & 0.08 & $\sim 10^6$ & \cite{Antheaume2008,Bedon2014} \\ 
        Sandia 17 m \cite{Worstell1978} & Darrieus  & 2 & 0.06 & $\sim 10^7$ & \cite{Para1988,Orlandi2015,Bedon2014} \\ 
        Sandia 34 m \cite{Ashwill1992} & Darrieus  & 2 & 0.05 & $\sim 10^7$ & \cite{Liu1992,Murray2011,Bedon2014}  \\ 
        Strickland \emph{et al.} \cite{Strickland1981} & H & 1--3 & 0.15 & $\sim 10^5$ & \cite{Ponta2001,Scheurich2011b} \\ 
        Laneville and Vitecoq \cite{Laneville1986} & H & 2 & 0.13 & $\sim 10^6$ & \cite{Amet2009} \\ 
        Howell \emph{et al.} \cite{Howell2010} & H & 2--3 & 0.33 & $\sim 10^5$ & \cite{Joo2015} \\ 
        Mertens \cite{Mertens2003} & H & 2 & 0.21 & $\sim 10^5$ & \cite{Orlandi2015} \\ 
        VAWT 850 \cite{Mays1990} & Tapered H & 2 & 0.05 & $\sim 10^7$ & \cite{Murray2011} \\ 
        UNH-RVAT \cite{Bachant2014-RVAT-baseline} & H & 3 & 0.28 & $\sim 10^6$ & \cite{Michelen2014} \\
        RM2 (present) & Tapered H & 3 & 0.07--0.12 & $\sim 10^6$ & 
    \end{tabular}

@@

In general, there have been more experiments done with low $c/R$ rotors. These
rotors are easier to model, since unsteady dynamic effects are less influential
on the overall performance \cite{Strickland1981}. This is apparent when
examining the effectiveness of numerical models that rely on static foil
coefficient input data, e.g., streamtube and vortex models, which are most
applicable for $c/R \leq 0.1$. For example, Bedon \emph{et al.}~\cite{Bedon2014}
used a double multiple streamtube momentum model without dynamic stall
corrections to evaluate the effectiveness of various foil coefficient databases
against the Sandia Darrieus turbine experimental data. Despite using such a
simple model, performance predictions were quite accurate in most conditions
except at low tip speed ratio for the 2 m turbine, which had the highest $c/R$
of all the Sandia rotors, making the dynamic effects more important. This
highlights the need for more validation data for higher solidity rotors to
ensure numerical models are robust enough to explore unique cross-flow turbine
designs, especially as the MHK concepts mature.

The UNH-RVAT, for which the performance, near-wake, and Reynolds number
dependence of was investigated using the same experimental setup as the study
here \cite{Bachant2015-JoT, Bachant2016-Energies}, was an H-rotor of 1 m height
and 1 m diameter. The UNH-RVAT experimental datasets are also openly available
\cite{Bachant2014-RVAT-baseline, Bachant2016-RVAT-Re-dep}, and provide an
interesting comparison for the near-wake dynamics of a rotor similar in size to
the 1:6 scale RM2, but with non-tapered blades and a high solidity $c/R = 0.28$.
